Static and Dynamic Bethe-Salpeter Equations in the $T$-Matrix Approximation

### 2022-02-12 Manuscript/Tmatrix.tex
minor corrections

One of the key features of the $T$-matrix approximation is its exactness up to the second order thanks to the inclusion of second-order exchange diagrams.
This class of diagrams, which are particularly important in molecular systems and explain the improvement brought by the second-order screened exchange (SOSEX) correction applied to $GW$, \cite{Romaniello_2009a,Ren_2015,Loos_2018b} are well known to be missing in the $GW$ approximation.

@@

The non-Hermitian pp-RPA problem reads \cite{Schuck_Book,vanAggelen_2013,Peng_2013,Scuseria_2013,Yang_2013,Yang_2013b,vanAggelen_2014,Yang_2014a,Zhang_2015,Zhang_2016}
\begin{equation}
\label{eq:LR-RPA}
	\begin{pmatrix}
		\bA{}{\ppRPA}			&	\bB{}{\ppRPA}	\\
		-\T{(\bB{}{\ppRPA})}	&	-\bC{}{\ppRPA}	\\
	\end{pmatrix}
	\cdot
	\begin{pmatrix}
		\bX{n}{N\pm2}	\\
		\bY{n}{N\pm2}	\\
	\end{pmatrix}
	=
	\Om{n}{N\pm2}
	\begin{pmatrix}
		\bX{n}{N\pm2}	\\
		\bY{n}{N\pm2}	\\
	\end{pmatrix}
\end{equation}
where the elements of the various matrices are defined as
\begin{subequations}
\begin{align}
	A_{ab,cd}^{\ppRPA} & = \delta_{ab} \delta_{cd} (\e{a}{} + \e{b}{}) + \mel{ab}{}{cd}
	\\ 
	B_{ab,ij}^{\ppRPA} & = \mel{ab}{}{ij}
	\\ 
	C_{ij,kl}^{\ppRPA} & = - \delta_{ik} \delta_{jl} (\e{i}{} + \e{j}{}) +\mel{ij}{}{kl}
\end{align}
\end{subequations}
and 
\begin{equation}
	\mel{pq}{}{rs} = \braket{pq}{rs} - \braket{pq}{sr}
\end{equation}
are two-electron integrals in the spinorbital basis, \ie,
\begin{equation}
	\braket{pq}{rs} = \iint \SO{p}(\bx_1) \SO{q}(\bx_2) \frac{1}{\abs{\br_1 - \br_2}} \SO{r}(\bx_1) \SO{s}(\bx_2)  d\bx_1 d\bx_2
\end{equation}
The pp-RPA problem yields, in the absence of instabilities (which should not appear in Coulombic systems with repulsive interactions only \cite{Scuseria_2013}), $\Nvir(\Nvir-1)/2$ positive eigenvalues $\Om{n}{N+2}$ and $\Nocc(\Nocc-1)/2$ negative eigenvalues $\Om{n}{N-2}$, which  correspond respectively to double attachments and double detachments.
The pp-RPA correlation energies is given by \cite{Peng_2013,Scuseria_2013}
\begin{equation}
\begin{split}
	\Ec^\ppRPA 
	& = + \sum_n \Om{n}{N+2} - \Tr(\bA{}{\ppRPA}) 
	\\
	& = - \sum_n \Om{n}{N-2} - \Tr(\bC{}{\ppRPA})
\end{split}
\end{equation}

@@

Within the (perturbative) one-shot $GT$ scheme (labeled as $G_0T_0$ in the following), the quasiparticle energies are obtained via linearization of the quasiparticle equation, \cite{Strinati_1980,Hybertsen_1985a,Hybertsen_1986,Godby_1988,Linden_1988,Northrup_1991,Blase_1994,Rohlfing_1995,Shishkin_2007} \ie,
\begin{equation}
\label{eq:G0T0}
	\e{p}{G_0T_0} = \e{p}{\HF} + Z_p \Sigma^{GT}_{pp}(\e{p}{\HF})
\end{equation}
where we have assumed a HF starting point and
\begin{equation}
	Z_p = \qty[ 1 - \eval{\pdv{\Sigma^{T}_{pp}(\omega)}{\omega}}_{\omega = \e{p}{\HF}} ]^{-1}
\end{equation}
is the renormalization factor.
Other levels of self-consistency can be considered like the partially self-consistent \textit{``eigenvalue''} $GT$ (ev$GT$)  \cite{Hybertsen_1986,Shishkin_2007,Blase_2011,Faber_2011,Rangel_2016,Gui_2018} or the quasiparticle self-consistent $GT$ (qs$GT$) \cite{Faleev_2004,vanSchilfgaarde_2006,Kotani_2007,Ke_2011,Kaplan_2016} schemes.

%%%%%%%%%%%%%%%%%%%%%%
\section{Neutral excitations}
\label{sec:neutral}
%%%%%%%%%%%%%%%%%%%%%%
Like the one-body Green's function is the pillar of the $GW$ and $GT$ approximations, the two-body Green's function $G_2$ is the key quantity of the Bethe-Salpeter equation (BSE) formalism of many-body perturbation theory \cite{Salpeter_1951,Strinati_1988,Blase_2018,Blase_2020} via its link with the two-body correlation function $L$ which satisfies the self-consistent BSE 
\begin{multline}
	 iL(1,2;1',2') = L_0(1,2;1',2') 
	 \\
	 + \int L_0(1,4;1',3) \Xi(3,5;4,6) L(6,2;5,2') d3d4d5d6
\end{multline}
where
\begin{subequations}
\begin{align}
\label{eq:L0}
	 iL_0(1,4;1',3) & = G(1,3) G(4,1')
	 \\
\label{eq:L}
	 iL(1,2;1',2') & = -G_2(1,2;1',2') + G(1,1')G(2,2')
\end{align}
\end{subequations}
and 
\begin{equation}
\label{eq:Xi}
	\Xi(3,5;4,6) = i \fdv{\Sigma(3,4)}{G(6,5)}
\end{equation}
is the so-called BSE kernel that takes into account the self-consistent variation of $\Sigma$ with respect to the variation of $G$.
By taking into account the interaction of the excited electron and its hole left behind (the infamous excitonic effect), the BSE is able to faithfully model (neutral) optical excitations as measured by absorption spectroscopy. 
The moderate cost of the BSE [which scales as $\order*{K^4}$ in its standard implementation] and its all-round accuracy are the main reasons behind its growing popularity in the molecular electronic structure community. \cite{Rohlfing_1999a,Horst_1999,Puschnig_2002,Tiago_2003,Boulanger_2014,Jacquemin_2015a,Bruneval_2015,Jacquemin_2015b,Hirose_2015,Jacquemin_2017a,Jacquemin_2017b,Rangel_2017,Krause_2017,Gui_2018,Blase_2018,Liu_2020,Blase_2020,Holzer_2018a,Holzer_2018b,Loos_2020e,Loos_2021}

@@

Based on Rayleigh-Schr\"odinger perturbation theory, the non-linear eigenproblem \eqref{eq:dBSE} can be split as a zeroth-order static reference and a first-order dynamic perturbation, such that
\begin{multline}
\label{eq:LR-PT}
	\begin{pmatrix}
		\bA{}{\dBSE}(\Om{S}{})		&	\bB{}{\BSE}(\Om{S}{})	\\
		-\bB{}{\BSE}(-\Om{S}{})	&	-\bA{}{\BSE}(-\Om{S}{})	\\
	\end{pmatrix}
	\\
	=
	\begin{pmatrix}
		\bA{}{\BSE}	&	\bB{}{\BSE}	
		\\
		-\bB{}{\BSE}	&	-\bA{}{\BSE}	
		\\
	\end{pmatrix}
	+
	\begin{pmatrix}
		\bA{}{(1)}(\Om{S}{})		&	\bO	\\
		\bO	&	-\bA{}{(1)}(-\Om{S}{})	\\
	\end{pmatrix}
\end{multline}
with
\begin{equation}
	\label{eq:BSE-A1}
	A_{ia,jb}^{(1)}(\omega) = \widetilde{T}_{ib,aj}(\omega) - T_{ib,aj}(\omega = 0)
\end{equation}
As usual, one can naturally expand the $S$th BSE excitation energy and its corresponding eigenvector as
\begin{subequations}
\begin{gather}
	\Om{S}{} = \Om{S}{\BSE} + \Om{S}{(1)} + \ldots,
	\\
	\begin{pmatrix}
		\bX{S}{}	\\
		\bY{S}{}	\\
	\end{pmatrix}
	= 
	\begin{pmatrix}
		\bX{S}{\BSE}	\\
		\bY{S}{\BSE}	\\
	\end{pmatrix}
	+
	\begin{pmatrix}
		\bX{S}{(1)}	\\
		\bY{S}{(1)}	\\
	\end{pmatrix}
	+ \ldots
\end{gather}
\end{subequations}
Solving the static BSE [see Eq.~\eqref{eq:BSE}] yields the (zeroth-order) static $\Om{S}{\BSE}$ excitation energies and their corresponding eigenvectors $\bX{S}{\BSE}$ and $\bY{S}{\BSE}$. 
The first-order correction to the $S$th excitation energy is, within the dynamical Tamm-Dancoff approximation,
\begin{equation}
\label{eq:Om1-TDA}
	\Om{S}{(1)} = \T{(\bX{S}{\BSE})} \cdot \bA{}{(1)}(\Om{S}{\BSE}) \cdot \bX{S}{\BSE}
\end{equation}
This correction can be renormalized by computing, at no extra cost, the renormalization factor which reads
\begin{equation}
\label{eq:Z}
	\zeta_{S} = \qty[ 1 - \T{(\bX{S}{\BSE})} \cdot \left. \pdv{\bA{}{(1)}(\Om{S}{})}{\Om{S}{}} \right|_{\Om{S}{} = \Om{S}{\BSE}} \cdot \bX{S}{\BSE} ]^{-1}
\end{equation}
This yields our final expression for the dynamically-corrected BSE excitation energies:
\begin{equation}
	\Om{S}{\text{dyn}} = \Om{S}{\text{stat}} + \Delta\Om{S}{\text{dyn}} = \Om{S}{\BSE} + \zeta_{S} \Om{S}{(1)}
\end{equation}
Note again that the present perturbative scheme does not allow to access double excitations as only excitations calculated within the static approach can be dynamically corrected.

>>>>

One of the key features of the $T$-matrix approximation is its exactness up to the second order thanks to the inclusion of second-order exchange diagrams.
This class of diagrams, which are particularly important in few-electron molecular systems \cite{Casida_1991,Hirata_2015,Hirata_2017} and explain the improvement brought by the second-order screened exchange (SOSEX) correction applied to $GW$, \cite{Romaniello_2009a,Ren_2015,Loos_2018b} are well known to be missing in the $GW$ approximation.

@@

The non-Hermitian pp-RPA problem reads \cite{Schuck_Book,vanAggelen_2013,Peng_2013,Scuseria_2013,Yang_2013,Yang_2013b,vanAggelen_2014,Yang_2014a,Zhang_2015,Zhang_2016}
\begin{equation}
\label{eq:LR-RPA}
	\begin{pmatrix}
		\bA{}{\ppRPA}			&	\bB{}{\ppRPA}	\\
		-\T{(\bB{}{\ppRPA})}	&	-\bC{}{\ppRPA}	\\
	\end{pmatrix}
	\cdot
	\begin{pmatrix}
		\bX{n}{N\pm2}	\\
		\bY{n}{N\pm2}	\\
	\end{pmatrix}
	=
	\Om{n}{N\pm2}
	\begin{pmatrix}
		\bX{n}{N\pm2}	\\
		\bY{n}{N\pm2}	\\
	\end{pmatrix}
\end{equation}
where the elements of the various matrices are defined as
\begin{subequations}
\begin{align}
	A_{ab,cd}^{\ppRPA} & = \delta_{ab} \delta_{cd} (\e{a}{} + \e{b}{}) + \mel{ab}{}{cd}
	\\ 
	B_{ab,ij}^{\ppRPA} & = \mel{ab}{}{ij}
	\\ 
	C_{ij,kl}^{\ppRPA} & = - \delta_{ik} \delta_{jl} (\e{i}{} + \e{j}{}) +\mel{ij}{}{kl}
\end{align}
\end{subequations}
and 
\begin{equation}
	\mel{pq}{}{rs} = \braket{pq}{rs} - \braket{pq}{sr}
\end{equation}
are two-electron integrals in the spinorbital basis, \ie,
\begin{equation}
	\braket{pq}{rs} = \iint \SO{p}(\bx_1) \SO{q}(\bx_2) \frac{1}{\abs{\br_1 - \br_2}} \SO{r}(\bx_1) \SO{s}(\bx_2)  d\bx_1 d\bx_2
\end{equation}
The pp-RPA problem yields, in the absence of instabilities (which should not appear in Coulombic systems with repulsive interactions only \cite{Scuseria_2013}), $\Nvir(\Nvir-1)/2$ positive eigenvalues $\Om{n}{N+2}$ and $\Nocc(\Nocc-1)/2$ negative eigenvalues $\Om{n}{N-2}$, which  correspond respectively to double attachments and double detachments.
The pp-RPA correlation energy is given by \cite{Peng_2013,Scuseria_2013}
\begin{equation}
\begin{split}
	\Ec^\ppRPA 
	& = + \sum_n \Om{n}{N+2} - \Tr(\bA{}{\ppRPA}) 
	\\
	& = - \sum_n \Om{n}{N-2} - \Tr(\bC{}{\ppRPA})
\end{split}
\end{equation}

@@

Within the (perturbative) one-shot $GT$ scheme (labeled as $G_0T_0$ in the following), the quasiparticle energies are obtained via linearization of the quasiparticle equation, \cite{Strinati_1980,Hybertsen_1985a,Hybertsen_1986,Godby_1988,Linden_1988,Northrup_1991,Blase_1994,Rohlfing_1995,Shishkin_2007} \ie,
\begin{equation}
\label{eq:G0T0}
	\e{p}{G_0T_0} = \e{p}{\HF} + Z_p \Sigma^{GT}_{pp}(\e{p}{\HF})
\end{equation}
where we have assumed a HF starting point and
\begin{equation}
	Z_p = \qty[ 1 - \eval{\pdv{\Sigma^{T}_{pp}(\omega)}{\omega}}_{\omega = \e{p}{\HF}} ]^{-1}
\end{equation}
is the renormalization factor or weight of the quasiparticle solution.
Other levels of (partial) self-consistency can be considered like the \textit{``eigenvalue''} self-consistent $GT$ (ev$GT$)  \cite{Hybertsen_1986,Shishkin_2007,Blase_2011,Faber_2011,Rangel_2016,Gui_2018} or the quasiparticle self-consistent $GT$ (qs$GT$) \cite{Faleev_2004,vanSchilfgaarde_2006,Kotani_2007,Ke_2011,Kaplan_2016} schemes.

%%%%%%%%%%%%%%%%%%%%%%
\section{Neutral excitations}
\label{sec:neutral}
%%%%%%%%%%%%%%%%%%%%%%
Like the one-body Green's function is the pillar of the $GW$ and $GT$ approximations, the two-body Green's function $G_2$ is the central quantity of the Bethe-Salpeter equation (BSE) formalism of many-body perturbation theory \cite{Salpeter_1951,Strinati_1988,Blase_2018,Blase_2020} via its link with the two-body correlation function $L$ which satisfies the self-consistent BSE 
\begin{multline}
	 iL(1,2;1',2') = L_0(1,2;1',2') 
	 \\
	 + \int L_0(1,4;1',3) \Xi(3,5;4,6) L(6,2;5,2') d3d4d5d6
\end{multline}
where
\begin{subequations}
\begin{align}
\label{eq:L0}
	 iL_0(1,4;1',3) & = G(1,3) G(4,1')
	 \\
\label{eq:L}
	 iL(1,2;1',2') & = -G_2(1,2;1',2') + G(1,1')G(2,2')
\end{align}
\end{subequations}
and 
\begin{equation}
\label{eq:Xi}
	\Xi(3,5;4,6) = i \fdv{\Sigma(3,4)}{G(6,5)}
\end{equation}
is the so-called BSE kernel that takes into account the self-consistent variation of $\Sigma$ with respect to the variation of $G$.
By taking into account the interaction of the excited electron and its hole left behind (the infamous excitonic effect), the BSE is able to faithfully model (neutral) optical excitations as measured by absorption spectroscopy. 
The moderate cost of the BSE [which scales as $\order*{K^4}$ in its standard implementation] and its all-round accuracy are the main reasons behind its growing popularity in the molecular electronic structure community. \cite{Rohlfing_1999a,Horst_1999,Puschnig_2002,Tiago_2003,Boulanger_2014,Jacquemin_2015a,Bruneval_2015,Jacquemin_2015b,Hirose_2015,Jacquemin_2017a,Jacquemin_2017b,Rangel_2017,Krause_2017,Gui_2018,Blase_2018,Liu_2020,Blase_2020,Holzer_2018a,Holzer_2018b,Loos_2020e,Loos_2021}

@@

Based on Rayleigh-Schr\"odinger perturbation theory, the non-linear eigenproblem \eqref{eq:dBSE} can be split as a zeroth-order static reference and a first-order dynamic perturbation, such that
\begin{multline}
\label{eq:LR-PT}
	\begin{pmatrix}
		\bA{}{\dBSE}(\Om{S}{})		&	\bB{}{\BSE}(\Om{S}{})	\\
		-\bB{}{\BSE}(-\Om{S}{})	&	-\bA{}{\BSE}(-\Om{S}{})	\\
	\end{pmatrix}
	\\
	=
	\begin{pmatrix}
		\bA{}{\BSE}	&	\bB{}{\BSE}	
		\\
		-\bB{}{\BSE}	&	-\bA{}{\BSE}	
		\\
	\end{pmatrix}
	+
	\begin{pmatrix}
		\bA{}{(1)}(\Om{S}{})		&	\bO	\\
		\bO	&	-\bA{}{(1)}(-\Om{S}{})	\\
	\end{pmatrix}
\end{multline}
with
\begin{equation}
	\label{eq:BSE-A1}
	A_{ia,jb}^{(1)}(\omega) = \widetilde{T}_{ib,aj}(\omega) - T_{ib,aj}(\omega = 0)
\end{equation}
As usual, one can naturally expand the $S$th BSE excitation energy and its corresponding eigenvector as
\begin{subequations}
\begin{gather}
	\Om{S}{} = \Om{S}{\BSE} + \Om{S}{(1)} + \ldots,
	\\
	\begin{pmatrix}
		\bX{S}{}	\\
		\bY{S}{}	\\
	\end{pmatrix}
	= 
	\begin{pmatrix}
		\bX{S}{\BSE}	\\
		\bY{S}{\BSE}	\\
	\end{pmatrix}
	+
	\begin{pmatrix}
		\bX{S}{(1)}	\\
		\bY{S}{(1)}	\\
	\end{pmatrix}
	+ \ldots
\end{gather}
\end{subequations}
Solving the static BSE [see Eq.~\eqref{eq:BSE}] yields the (zeroth-order) static $\Om{S}{\BSE}$ excitation energies and their corresponding eigenvectors $\bX{S}{\BSE}$ and $\bY{S}{\BSE}$. 
The first-order correction to the $S$th excitation energy is, within the dTDA,
\begin{equation}
\label{eq:Om1-TDA}
	\Om{S}{(1)} = \T{(\bX{S}{\BSE})} \cdot \bA{}{(1)}(\Om{S}{\BSE}) \cdot \bX{S}{\BSE}
\end{equation}
This correction can be renormalized by computing, at no extra cost, the renormalization factor which reads
\begin{equation}
\label{eq:Z}
	\zeta_{S} = \qty[ 1 - \T{(\bX{S}{\BSE})} \cdot \left. \pdv{\bA{}{(1)}(\Om{S}{})}{\Om{S}{}} \right|_{\Om{S}{} = \Om{S}{\BSE}} \cdot \bX{S}{\BSE} ]^{-1}
\end{equation}
This yields our final expression for the dynamically-corrected BSE excitation energies:
\begin{equation}
	\Om{S}{\text{dyn}} = \Om{S}{\text{stat}} + \Delta\Om{S}{\text{dyn}} = \Om{S}{\BSE} + \zeta_{S} \Om{S}{(1)}
\end{equation}
Note again that the present perturbative scheme does not allow to access double excitations as only excitations calculated within the static approach can be dynamically corrected.
